$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-09-21 paper/paper.tex
Update paper.tex

A courier system consists of three roles:
a \sender{}, who wants to send a package;
a \receiver{}, who receives the package sent by the sender;
and a \courier{}, who carries the package from the sender to the receiver.
Our courier system is defined by four \picalc{} processes in parallel inhabiting the above three roles:
we have two \sender{} processes, $S_1$ and $S_2$, sending data over channels $x$ and $y$, respectively;
one \receiver{} process, $R$, which receives the data sent from each of the senders--hence receives twice;
and a \courier{} process $C$, which synchronises communication among $S_1$, $S_2$ and $R$.
The \courier{} process $C$ first receives data from $S_1$ and $S_2$ on its input channels $x$ and $y$, respectively, and then sends the received data to $R$ along its output channel $z$.
A graphical representation of the courier system is given in \autoref{fig:example-process}.

We will now define the courier system as \picalc{} processes.
In our framework, the syntax of \picalc{} processes adopts \emph{de Bruijn indices} \cite{deBruijn1972, Dybjer1994} for channel names and variables and is introduced in detail in \autoref{syntax}.
In short, for the purposes of this example:
process $\new{} P$ creates a new channel and makes it available at index $0$ in the continuation process $P$;
$\comp{P}{Q}$ composes processes $P$ and $Q$ in parallel; process $\recv{x}{}P$ receives data along channel $x$ and makes that data available at index $0$ in the continuation process $P$;
process $\send{x}{y}{P}$ sends variable $y$ over channel $x$ and continues as process $P$.
Both variable references $\Var_n$ and processes $\Process_n$ are indexed by the number $n$ of free variables they can refer to, for any natural number $n$.

For the sake of clarity, we will define the four subprocesses in our courier system separately and parametrise them by the channels on which they operate.

@@

    \draw[->] (send-x) -- node[left] {$x$} (sync);
    \draw[->] (send-y) -- node[right] {$y$} (sync);
    \draw[->] (sync.35) -- node[right] {$z$} (recv-z.240);
    \draw[->] (sync) -- (recv-z);
  \end{tikzpicture}
  \caption{The courier system: two \sender{} $S_1$ and $S_2$, one \receiver{} $R$ and one \courier{} $C$.}
  \label{fig:example-process}
\end{figure}

Each of the roles \sender{}, \receiver{} and \courier{} is defined as a function respectively, $\func{send}, \func{recv}$ and $\func{carry}$ with signature $\Var_n \to \Process_n$.

The \sender{} role is defined below.
The \sender{} creates a new channel $\new{} \suc c$ denoting the data to be sent, and sends it over channel $c$, and then terminates, denoted by $\P0$.
\begin{flalign*}
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{(\suc c)}{\constr{0}} \PO
\end{flalign*}
Each \sender{} process $S_1$ and $S_2$ will  be an instantiation of $\func{send} \; c$.
In particular
$S_1$ will be defined as $\func{send} \; x$ and $S_2$ as $\func{send} \; y$, where data is sent over channels $x$ and $y$, respectively (see \autoref{fig:example-process}).

The \receiver{} role is defined below, where $\func{recv} \; c \;$ receives data \emph{twice} on a channel $c$ and then terminates with $\PO$.
\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} \recv{(\suc c)} \PO
\end{flalign*}
The \receiver{} process $R$ will be instantiated as $\func{recv} \; z$, where $z$ is the channel used to receive data from the \courier{} (see \autoref{fig:example-process}).

The \courier{} role is defined below.
\begin{flalign*}
  & \func{carry} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
  & \func{carry} \; in0 \; in1 \; out \; &&= \\
  & && \recv{in0}{} \\
  & && \recv{(\suc \; in1)}{} \\
  & && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
  & && \send{(\suc \suc \; out)}{\constr{0}} \PO
\end{flalign*}
The \courier{} process $C$ will be instantiated as $\func{carry} \; x \; y \; z$, which sequentially receives on the two input channels $in0$ and $in1$, and then outputs the two pieces of received data on the output channel $out$. Here channels $x,y,z$ instantiate respectively $in0,in1,out$.

@@

In order to type the courier system we will use all three typing disciplines: linear, graded and shared typing.
We will continue this example and provide the typing derivations in \autoref{leftover-typing}.
We use \textbf{shared types} to type the data that is being sent from the \sender{} processes to the \receiver{};
we use \textbf{linear types} for the usage annotations of [FILL IN, NOT CLEAR BEFORE];
and finally, we use \textbf{graded types} for the usage annotations of the communication channels $x$, $y$, and $z$: the \sender{} processes $\func{send} \; x$ and $\func{send} \; y$ each consume $1$ output multiplicity and the \receiver{} process $\func{recv} \; z$ consumes $2$ input multiplicities, as it receives twice on channel $z$.
Dually, the \courier{} process $\func{carry} \; x \; y \; z$ consumes one input multiplicity on $x$ and one input multiplicity on $y$, and two output multiplicities on $z$.
\end{example}

@@

   \item \textbf{Semantics}: we provide an operational semantics for the \picalc{}, prior to any typing (\autoref{semantics}).
   This operational semantics is defined as a reduction relation on processes (\autoref{operational-semantics}).
   The reduction relation tracks at the type level the channel on which communication occurs. This information is later used to state the subject reduction theorem --- aka type preservation (\autoref{thm:subject-reduction}).
   The reduction relation is defined modulo \emph{structural congruence} (\autoref{structural-congruence}) --- a relation defined on processes that acts as a quotient type to remove unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
 \end{itemize}

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double inference lines and index-free synonyms (\textsc{Nat}) as rule names (for ease of reference).
Constructors ($\constr{0}$ and $\suc$) are given as inference rules.
We maintain a close correspondence between the definitions presented in this paper and our mechanised definitions in Agda: premises become argument types and conclusions return types.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned in type definitions but omitted by constructors.

@@

\begin{example}[$\func{fromRaw}$ and $\func{toRaw}$]
  \begin{figure}[t]
    \begin{alignat*}{7}
      &P = (\new{x} ) && (\comp {\recv{x}{x} && \send{x}{z} && \PO} {(\new{y}) && (\send{x}{y} && \recv{y}{y} && \PO)}) \\
      &Q = \new{} && (\comp {\recv{0}{} && \send{0}{2} && \PO} {\new{} && (\send{1}{0} && \recv{0}{} && \PO)}) \\
      &R = (\new{x^0} ) && (\comp {\recv{x^0}{x^1} && \send{x^1}{z^0} && \PO} {(\new{y^0}) && (\send{x^0}{y^0} && \recv{y^0}{y^1} && \PO)})
    \end{alignat*}
    \caption{From names to de Bruijn indices}
    \label{fig:conversion}
    \end{figure}
  We illustrate the conversion functions from names to de Bruijn indices $\func{fromRaw}$ and back $\func{toRaw}$ with three processes $P,Q,R$ given in \autoref{fig:conversion}.

  Process $P$ uses names $x,y,z$ and it is translated via the conversion function $\func{fromRaw}$ into process $Q$, which uses de Bruijn indices.
  Process $Q$ is then translated via $\func{toRaw}$ into process $R$, which follows the \emph{Barendregt convention}\footnote{The Barendregt variable convention states that all bound variables/names in a process are distinct among each other and from the free variables/names.} and is $\alpha$-equivalent to the original process $P$.
\end{example}

@@

The first three rules ($\constr{comp}-$*) state associativity, symmetry, and $\PO$ as being the neutral element of parallel composition, respectively.
The last three ($\constr{scope}-$*) state garbage collection, scope extrusion and commutativity of restrictions, respectively.

In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof asserting that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$ (see \autoref{app:definitions}).
The function $\func{lower}_i \; Q \; uQ$ traverses $Q$ decrementing every index greater than $i$.
In $\constr{scope-comm}$ the function $\func{exchange}_i \; P$ traverses $P$ (of type $\allowbreak \Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

We lift the relation \textsc{StrucCong} $\eqeq{}$ and close it under equivalence and congruence in \textsc{Equals} $\eq{}$ as shown in \autoref{fig:struct-cong1}.
The $\func{cong-}$* rules define structural congruence under a context $\mathcal{C}[\cdot]$ \cite{Sangio01}, respectively restriction, composition, input and output.
The remaining three rules close structural congruence under reflexivity, symmetry and transitivity. 
The transitivity rule makes it tricky to prove predicates by induction on \textsc{Equals}: Agda cannot immediately see that such doubly recursive calls terminate.
In order to remedy this, we model the inductive structure syntactically and index \textsc{Equals} by a type \textsc{Rec}.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{thm:subject-reduction}).
We distinguish between channels that are created inside of the process ($\constr{internal}$) and channels that are created outside ($\constr{external}\ i$), where $i$ is the index of the channel variable).

In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After this substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (we apply \autoref{lm:substitution-unused} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

@@

In $\constr{res}$, we use $\func{dec}$ to decrement the index of channel $c$ as we wrap processes $P$ and $Q$ inside a binder.
This $\func{dec}$ function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

@@

\autoref{fig:algebra-instances} sketches the implementation of linear, graded and shared types as instances of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivation}.

>>>>

A courier system consists of three \emph{roles}:
a \sender{}, who wants to send a package;
a \receiver{}, who receives the package sent by the sender;
and a \courier{}, who carries the package from the sender to the receiver.

Our courier system (given in \autoref{fig:example-process}) is defined by four \picalc{} processes in parallel instantiating the above three roles:
we have two \sender{} processes, $S_1$ and $S_2$, sending data over channels $x$ and $y$, respectively;
one \receiver{} process, $R$, which receives over channel $z$ the data sent from each of the senders--hence receives twice;
and a \courier{} process $C$, which synchronises communication among $S_1$, $S_2$ and $R$.
The \courier{} process $C$ first receives data from $S_1$ and $S_2$ on its input channels $x$ and $y$, respectively, and then sends the received data to $R$ along its output channel $z$.

We will now define the courier system as \picalc{} processes.
In our work, the syntax of \picalc{} processes adopts \emph{de Bruijn indices} \cite{deBruijn1972} for channel names and variables (details in \autoref{syntax}).
In short, for the purposes of this example:
process $\new{} P$ creates a new channel and makes it available at index $0$ in the continuation process $P$;
$\comp{P}{Q}$ composes processes $P$ and $Q$ in parallel; process $\recv{x}{}P$ receives data along channel $x$ and makes that data available at index $0$ in the continuation process $P$;
process $\send{x}{y}{P}$ sends variable $y$ over channel $x$ and continues as process $P$.
Both variable references $\Var_n$ and processes $\Process_n$ are indexed by the number $n$ of free variables they can refer to, for any natural number $n$.

@@

    \draw[->] (send-x) -- node[left] {$x$} (sync);
    \draw[->] (send-y) -- node[right] {$y$} (sync);
    \draw[->] (sync.35) -- node[right] {$z$} (recv-z.240);
    \draw[->] (sync) -- (recv-z);
  \end{tikzpicture}
  \caption{The courier system: two \sender{} $S_1$ and $S_2$, one \receiver{} $R$ and one \courier{} $C$.}
  \label{fig:example-process}
\end{figure}
For the sake of clarity, we will define the four subprocesses in our courier system separately and parametrise them by the channels on which they operate.
Each roles \sender{}, \receiver{} and \courier{} is defined as a function respectively, $\func{send}, \func{recv}$ and $\func{carry}$ with signature $\Var_n \to \Process_n$.

The \sender{} \emph{role} is defined below.
The \sender{} creates a new channel $\new{} \suc c$ denoting the data to be sent, and sends it over channel $c$, and then terminates, denoted by process $\PO$.
\begin{flalign*}
  & \func{send} \; : \; \Var_n \to \Process_n & \\
  & \func{send} \; c = \new{} \send{(\suc c)}{\constr{0}} \PO
\end{flalign*}
Each \sender{} \emph{process} $S_1$ and $S_2$ is an instantiation of $\func{send} \; c$.
In particular,
$S_1$ is defined as $\func{send} \; x$ and $S_2$ as $\func{send} \; y$, where data is sent over channels $x$ and $y$, respectively (see \autoref{fig:example-process}).

The \receiver{} role is defined below, where $\func{recv} \; c \;$ receives data \emph{twice} on a channel $c$ and then terminates with $\PO$.
\begin{flalign*}
  & \func{recv} \; : \; \Var_n \to \Process_n & \\
  & \func{recv} \; c \; = \recv{c}{} \recv{(\suc c)} \PO
\end{flalign*}
The \receiver{} process $R$ is defined as  $\func{recv} \; z$, with channel $z$ instantiating $c$, which is the channel used to receive data from \courier{} (see \autoref{fig:example-process}).

The \courier{} role is defined below.
The \courier{} process $C$ is defined as $\func{carry} \; x \; y \; z$.
It sequentially receives on the two input channels $x$ and $y$ -- instantiating $in0$ and $in1$, and then outputs the two pieces of received data on the output channel $z$ -- instantiating $out$ (see \autoref{fig:example-process}).
\begin{flalign*}
  & \func{carry} \; : \; \Var_n && \to \Var_n \to \Var_n \to \Process_n & \\
  & \func{carry} \; in0 \; in1 \; out \; &&= \\
  & && \recv{in0}{} \\
  & && \recv{(\suc \; in1)}{} \\
  & && \send{(\suc \suc \; out)}{\suc \constr{0}} \\
  & && \send{(\suc \suc \; out)}{\constr{0}} \PO
\end{flalign*}

@@

In order to type the courier system we will use all three typing disciplines: linear, graded and shared typing.
We will continue this example and provide the typing derivations in \autoref{leftover-typing}.
We use \textbf{linear types} for the \emph{usage annotations} of data and \textbf{shared types} for the \emph{data} itself that is being sent from the \sender{} processes to the \receiver{}.
We use \textbf{graded types} for the usage annotations of the communication channels $x$, $y$, and $z$: the \sender{} processes $\func{send} \; x$ and $\func{send} \; y$ each consume $1$ output multiplicity and the \receiver{} process $\func{recv} \; z$ consumes $2$ input multiplicities, as it receives twice on channel $z$.
Dually, the \courier{} process $\func{carry} \; x \; y \; z$ consumes one input multiplicity on $x$ and one input multiplicity on $y$, and two output multiplicities on $z$.
\end{example}

@@

   \item \textbf{Semantics}: we provide an operational semantics for the \picalc{}, prior to any typing (\autoref{semantics}).
   Thanks to our well-scoped grammar  (\autoref{syntax}), the operational semantics is defined on \emph{the totality of the syntax} as a reduction relation on processes (\autoref{operational-semantics}).
   The reduction relation tracks at the type level the channel on which communication occurs. This information is later used to state the subject reduction theorem --- aka type preservation (\autoref{thm:subject-reduction}).
   The reduction relation is defined modulo \emph{structural congruence} (\autoref{structural-congruence}) --- a relation defined on processes that acts as a quotient type to remove unnecessary syntactic minutiae introduced by the syntax of the \picalc{}.
 \end{itemize}

@@

\autoref{fig:notation} illustrates the notation used in this paper.
Data type definitions ($\type{\N}$) use double inference lines and index-free synonyms (\textsc{Nat}) as rule names for ease of reference.
Constructors ($\constr{0}$ and $\suc$) are given as inference rules.
We maintain a close correspondence between the definitions presented in this paper and our mechanised definitions in Agda: premises become argument types and conclusions return types.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned in type definitions but omitted by constructors.

@@

\begin{example}[$\func{fromRaw}$ and $\func{toRaw}$]
  \begin{figure}[t]
    \begin{alignat*}{7}
      &P = (\new{x} ) && (\comp {\recv{x}{x} && \send{x}{z} && \PO} {(\new{y}) && (\send{x}{y} && \recv{y}{y} && \PO)}) \\
      &Q = \new{} && (\comp {\recv{0}{} && \send{0}{2} && \PO} {\new{} && (\send{1}{0} && \recv{0}{} && \PO)}) \\
      &R = (\new{x^0} ) && (\comp {\recv{x^0}{x^1} && \send{x^1}{z^0} && \PO} {(\new{y^0}) && (\send{x^0}{y^0} && \recv{y^0}{y^1} && \PO)})
    \end{alignat*}
    \caption{From names to de Bruijn indices and back}
    \label{fig:conversion}
    \end{figure}
  We illustrate the conversion functions from names to de Bruijn indices $(\func{fromRaw}$) and back ($\func{toRaw}$) with three processes $P,Q,R$  in \autoref{fig:conversion}.

  Process $P$ uses names $x,y,z$ and is translated via the conversion function $\func{fromRaw}$ into process $Q$, which uses de Bruijn indices.
  Process $Q$ is then translated via $\func{toRaw}$ into process $R$, which follows the \emph{Barendregt convention}\footnote{The Barendregt variable convention states that all bound variables/names in a process are distinct among each other and from the free variables/names.} and is $\alpha$-equivalent to the original process $P$.
\end{example}

@@

The first three rules ($\constr{comp}-$*) state associativity, symmetry, and $\PO$ as being the neutral element of parallel composition, respectively.
The last three ($\constr{scope}-$*) state garbage collection, scope extrusion and commutativity of restrictions, respectively.
In $\constr{scope-ext}$ the type $\Unused_i \; Q$ is an inductive proof asserting that the variable index $i$ does not appear neither in the inputs nor in the outputs of $Q$ (see \autoref{app:definitions}).
The function $\func{lower}_i \; Q \; uQ$ traverses $Q$ decrementing every index greater than $i$.
In $\constr{scope-comm}$ the function $\func{exchange}_i \; P$ traverses $P$ (of type $\allowbreak \Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all the above, $i$ is incremented every time we go under a binder.

We lift the relation \textsc{StrucCong} $\eqeq{}$ and close it under equivalence and congruence in \textsc{Equals} $\eq{}$ as shown in \autoref{fig:struct-cong1}.
The $\constr{cong-}$* rules define structural congruence under a context $\mathcal{C}[\cdot]$ \cite{Sangio01}, respectively scope restriction, parallel composition, input and output.
The remaining three rules close structural congruence under reflexivity, symmetry and transitivity, respectively. 
The transitivity rule makes it tricky to prove predicates by induction on \textsc{Equals}: Agda cannot immediately see that such doubly recursive calls terminate.
In order to remedy this, we model the inductive structure syntactically and index \textsc{Equals} by a type \textsc{Rec}.

@@

The operational semantics of the \picalc{} is defined as a reduction relation \textsc{Reduces} $\reduce{c}$ indexed by the channel $c$ on which communication occurs (\autoref{fig:reduction}).
We keep track of channel $c$ so we can state subject reduction (\autoref{thm:subject-reduction}).

We distinguish between channels that are created inside of the process ($\constr{internal}$), and channels that are created outside ($\constr{external}\ i$), where $i$ is the index of the channel variable.
In rule $\constr{comm}$, parallel processes reduce when they communicate over a common channel with index ${i}$.
As a result of that communication, the continuation of the input process $P$ has all the references to its most immediate variable substituted with references to $\suc j$, the variable sent by the output process $\send{i}{j}{Q}$.
After this substitution, $\subst{P}{\suc j}{\constr{0}}$ is \emph{lowered} --- all variable references are decreased by one (we apply \autoref{lm:substitution-unused} to obtain a proof $\Unused_{\constr{0}} \; (\subst{P}{\suc j}{\constr{0}})$).
Reduction is closed under parallel composition (rule $\constr{par}$), restriction (rule $\constr{res}$) and structural congruence (rule $\constr{struct}$) 
--- notably, not under input nor output, as doing so would not preserve the sequencing of actions \cite{Sangio01}.

@@

Rule $\constr{res}$ uses $\func{dec}$ to decrement the index of channel $c$ as we wrap processes $P$ and $Q$ inside a binder.
This function saturates at $\constr{internal}$: wrapping an internally reducing process with a binder will result in an internally reducing process.

@@

\autoref{fig:algebra-instances} sketches the implementation of linear, graded and shared types as instances of our usage algebra.
Their use in typing derivations is illustrated in \autoref{example-derivation} (Continued).
